Model Recognition as Planning

### 2018-10-30 doc/icaps-19/model-recognition.tex
first icaps19 paper review to address new comment by Miquel

This paper formalizes the {\em model recognition} task where the object to recognize is not a goal but the {\em planning model} that is determining the behavior of the observed agent. Given a partially observed plan execution, and a set of possible planning models (models that share the same state variables but that update these variables with different action models); {\em model recognition} is the task of identifying the model in the set with the highest probability of producing/explaining the given observation.

@@

To better illustrate {\em model recognition}, imagine a robot in a $n\times n$ grid whose navigation is determined by the \strips\ model of Figure~\ref{fig:model-example}. According to this model the robot could increment its {\em x-coordinate} when {\tt\small $q0$} holds (i.e. at {\em even rows} if {\tt\small $q0$} holds initially) and decrement it when {\tt\small $q1$} holds (at {\em odd rows} if {\tt\small $q0$} holds initially). Apart from this particular navigation model, different action models could be defined within the same state variables (e.g. altering the way {\tt\small $q0$} and {\tt\small $q1$} are required and updated) and these models can determine different kinds of robot navigation. Given an observation of a plan execution (like the one illustrated at Figure~\ref{fig:grid-example}) {\em model recognition} would aim here to identify which navigation model produced/explains that observation, despite key information is unobserved (e.g. the applied actions or the value of state variables {\small\tt $q0$} and {\small\tt $q1$}).

{\em Model recognition} is of interest because once the planning model is recognized, then the model-based machinery for automated planning becomes applicable~\cite{ghallab2004automated}. In addition, it enables identifying different kinds of automata by observing their execution. It is well-known that diverse automata representations, like {\em finite state controllers}, {\em push-down automata}, {\em {\sc GOLOG} programs} or {\em reactive policies}, can be encoded as classical planning models~\cite{baier2007exploiting,Geffner:FSM:AAAI10,ivankovic2015optimal,segovia2017generating}.

@@

  (:action inc-y-even
    :parameters (?y1 ?y2)
    :precondition (and (ycoord ?y1) (next ?y1 ?y2) (q0))
    :effect (and (not (ycoord ?y1)) (ycoord ?y2)
                 (not (q0)) (q1)))

  (:action inc-y-odd
    :parameters (?y1 ?y2)
    :precondition (and (ycoord ?y1) (next ?y1 ?y2) (q1))
    :effect (and (not (ycoord ?y1)) (ycoord ?y2)
                 (not (q1)) (q0)))

  (:action dec-y-even
    :parameters (?y1 ?y2)
    :precondition (and (ycoord ?y1) (next ?y2 ?y1) (q0))
    :effect (and (not (ycoord ?y1)) (ycoord ?y2)
                 (not (q0)) (q1)))

  (:action dec-y-odd
    :parameters (?y1 ?y2)
    :precondition (and (ycoord ?y1) (next ?y2 ?y1) (q1))
    :effect (and (not (ycoord ?y1)) (ycoord ?y2)
                 (not (q1)) (q0)))
  \end{verbatim}           
  \end{tiny}  
 \caption{\small Example of a \strips\ action model (codded in PDDL) for robot navigation in a $n\times n$ grid.}
\label{fig:model-example}
\end{figure}

@@

\subsection{The observation model}
Given a classical planning problem $P=\tup{F,A,I,G}$ and a plan $\pi$ that solves $P$; the observation of the execution of $\pi$ on $P$ is $\mathcal{O}(\pi,P)=\tup{a_1^o,s_1^o \ldots , a_l^o, s_m^o}$, an interleaved combination of {\small $1\leq l\leq |\pi|$} observed actions and {\small $1\leq m\leq |\pi|$} observed states such that:
\begin{itemize}
\item Observed actions are {\em consistent} with $\pi$~\cite{ramirez2009plan}. This means that $\tup{a_1^o, \ldots, a_l^o}$ is a sub-sequence of the solution plan $\pi$.
\item Observed states are a sequence of partial states $\tup{s_1^o, \ldots, s_m^o}$ that is {\em consistent} with the sequence of states $\tup{s_0, s_1, \ldots, s_n}$ traversed by the execution of $\pi$ on $P$. 
\end{itemize}

On the one hand, the initial state $I$ is fully observed while the observed states in $\mathcal{O}(\pi,P)$ may be partial, i.e. the value of certain fluents in the intermediate states may be omitted (~$|s_i^o|\leq |F|$ for every $1\leq i\leq m$). On the other hand, the sequence of observed states $\tup{s_1^o, \ldots, s_m^o}$ in $\mathcal{O}(\pi,P)$ corresponds to the same sequence of states traversed by $\pi$ but certain states may also be omitted ($0\leq|s_i^o|$ for every $1\leq i\leq m$). This means that the transitions between two consecutive observed states in $\mathcal{O}(\pi,P)$ may require the execution of more than a single action ($\theta(s_i^o,\tup{a_1,\ldots,a_k})=s_{i+1}^o$, where ${\small k\geq 1}$ is unknown and unbound). Therefore we can conclude that having $\mathcal{O}(\pi,P)$ does not imply knowing the actual length of $\pi$.

\begin{definition}[$\Phi$-observation]
Given a subset of fluents $\Phi\subseteq F$ we say that $\mathcal{O}(\pi,P)$ is a $\Phi$-observation of the execution of $\pi$ on $P$ iff, for every ${\small 1\leq i\leq m}$, each observed state $s_i^o$ only contains fluents in $\Phi$.
\end{definition}

Figure~\ref{fig:grid-example} illustrates the six-state $\Phi$-observation \{{\tt\scriptsize<(xcoord 2)(ycoord 1)>, <(xcoord 3)(ycoord 1)>, <(xcoord 4)(ycoord 1)>, <(xcoord 5)(ycoord 1)>, <(xcoord 5)(ycoord 2)>, <(xcoord 4)(ycoord 2)>}\}. In this case $\Phi$ only contains fluents of the kind {\tt\small (xcoord ?v)} and {\tt\small (ycoord ?v)} so that the value of the remaining fluents, namely {\tt\small (next ?v1 ?v2)}, {\tt\small (q0)} and {\tt\small (q1)}, is unknown.

\section{Model Recognition}
\label{sec:recognition}
The {\em model recognition} task is a tuple $\tup{P,M,\mathcal{O}}$ where:
\begin{itemize}
\item $P=\tup{F,A[\cdot],I,G}$ is a {\em classical planning problem} s.t. the semantics of actions $a\in A[\cdot]$ is unknown (i.e. the corresponding $\rho$ and/or $\theta$ functions are undefined).
\item $M=\{\mathcal{M}_1,\ldots,\mathcal{M}_m\}$ is a finite non-empty {\em set of models} for the actions in $A[\cdot]$ s.t., each model in $\mathcal{M}\in M$, defines different function pairs $\tup{\rho,\theta}$ over the state variables $F$.
\item $\mathcal{O}(\pi,P)$ is an {\em observation} of the execution of an unknown solution plan $\pi$ for the planning problem $P$.
\end{itemize}

{\em Model recognition} can be understood as a {\em classification task} where each class is represented with a different planning model $\mathcal{M}\in M$ and the observed plan execution $\mathcal{O}(\pi,P)$, is the single example to classify. The planning model associated to each class acts as the corresponding {\em class prototype} and it summarizes any observation of a plan execution that could be synthesized with that model (i.e. all the examples that belong to that class).

We follow the {\em naive Bayes classifier} to assign a model $\mathcal{M}\in M$ to the given observation $\mathcal{O}(\pi,P)$ with respect to the expression:
\begin{align}
argmax_{\mathcal{M}\in M} P(\mathcal{O}|\mathcal{M})\times P(\mathcal{M}).
\end{align}
The {\em solution} to the {\em model recognition} task is the subset of models in $M$ that maximizes expression (1).

\subsection{Formulating the $P(\mathcal{O}|\mathcal{M})$ {\em likelihood}}
The $P(\mathcal{M})$ probability expresses whether one model is known to be a priori more likely than the others. If this probability is not given as input it is reasonable to assume that, {\em a priori}, all models are equiprobable. The challenge in our formulation for {\em model recognition} is the definition of the $P(\mathcal{O}|\mathcal{M})$ {\em likelihood} that expresses the probability of observing $\mathcal{O}(\pi,P)$ when $\mathcal{M}$ is the planning model.

\begin{figure}
  \begin{scriptsize}
  \begin{center}
	\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=2cm,semithick]
	  \node[state] (A)              {$\mathcal{M}$};
	  \node[state] (B) [right of=A] {$\mathcal{M}'$};
	  \node[state] (C) [right of=B] {$\tau$};
	  \node[state] (D) [right of=C] {$\mathcal{O}$};          
	  \path
(A) edge  node {} (B)
(B) edge  node {} (C)
(C) edge  node {} (D)
;
	\end{tikzpicture}
  \end{center}
  \end{scriptsize}  
 \caption{\small {\em Bayesian network} abstracting that model $\mathcal{M}$ can be {\em transformed} into a model $\mathcal{M}'$ that is able to produce a trajectory $\tau(\pi,P)$ that reaches the goals in $P$ and is consistent with the observation $\mathcal{O}(\pi,P)$.}
\label{fig:net}
\end{figure}

Our approach to formulate the $P(\mathcal{O}|\mathcal{M})$ {\em likelihood} is to assess the cost of {\em transforming} $\mathcal{M}$ into a model $\mathcal{M'}$ that can produce a trajectory $\tau(\pi,P)$ that that reaches the goals in $P$ and is consistent with the observation $\mathcal{O}(\pi,P)$. Figure~\ref{fig:net} shows the four-variable {\em Bayesian network} abstracting this process. Regarding this network we have the following formulation:
\begin{align}
 P(\mathcal{O}|\mathcal{M})=\sum_{\mathcal{M}'}\sum_{\tau} P(\mathcal{M'}|\mathcal{M})P(\tau|\mathcal{M'})P(\mathcal{O}|\tau),
\end{align}
where $\tau$ ranges over all the trajectories that can be synthesized with a model $\mathcal{M}'$ and $\mathcal{M}'$ ranges over all the models that can be generated {\em transforming} $\mathcal{M}$.

{\em Edit distances} are similarity metrics, traditionally computed over {\em strings} or {\em graphs}, and that has been proved successful for {\em pattern recognition}~\cite{masek1980faster,bunke1997relation}. We assess the cost of {\em transforming} a classical planning model $\mathcal{M}$ into a model $\mathcal{M'}$ formalizing and computing {\em edit distances} that, in this work, are referred to \strips\ planning models.

\section{Recognition of \strips\ models}
\label{sec:asPlanning}
Here we analyze the particular instantiation of the {\em model recognition} task where the semantics of actions (i.e. $\rho$ and $\theta$ functions) are specified with \strips\ action schema. We start formalizing the \strips\ schema and define the full space of possible \strips\ schema. Eventually, we introduce an {\em edit distance} for \strips\ schema to estimate the $P(\mathcal{O}|\mathcal{M})$ likelihoods for classical planning models.

\subsection{Well-defined \strips\ action schema}
\strips\ action schema provide a compact representation for specifying classical planning models. Figure~\ref{fig:model-example} shows six \strips\ action schema, codded in PDDL, that determine a particular kind of robot navigation in $n\times n$ grids.

{\em A \strips\ action schema} $\xi$ is defined by a list of {\em parameters} $pars(\xi)$, and three list of predicates (namely $pre(\xi)$, $del(\xi)$ and $add(\xi)$) that shape the kind of fluents that can appear in the {\em preconditions}, {\em negative effects} and {\em positive effects} of the actions induced from that schema.

We say that two \strips\ schemes $\xi$ and $\xi'$ are {\em comparable} iff $pars(\xi)=pars(\xi')$, both share the same list of parameters. For instance, we claim that the six action schema of Figure~\ref{fig:model-example} are {\em comparable} while, for example, the {\small\tt stack(?v1,?v2)} and {\small\tt pickup(?v1)} schemes from the four operator {\em blocksworld}~\cite{slaney2001blocks} are not. Last but not least, we say that two \strips\ models $\mathcal{M}$ and $\mathcal{M}'$ are {\em comparable} iff there exists a bijective function $\mathcal{M} \mapsto \mathcal{M}^*$ that maps every action schema $\xi\in\mathcal{M}$ to a comparable schema $\xi'\in\mathcal{M'}$ and vice versa.

Given a \strips\ action schema $\xi$, let us define an additional set of objects ($\Omega\cap\Omega_\xi=\emptyset$), that we denote as {\em variable names}, and that contains one variable name for each parameter in $pars(\xi)$, that is $\Omega_\xi=\{v_i\}_{i=1}^{|pars(\xi)|}$. For any of the six schema defined in Figure~\ref{fig:model-example}, $|pars(\xi)|=2$ so $\Omega_\xi=\{v_1,v_2\}$.

Given a \strips\ action schema $\xi$ and a set of {\em predicates} $\Psi$ that shape the propositional state variables. The set of FOL interpretations of $\Psi$ over the corresponding $\Omega_\xi$ objects ({\em variable names} for schema $\xi$), confines the elements that can appear in the $pre(\xi)$, $del(\xi)$ and $add(\xi)$ lists. We denote this set of FOL interpretations as ${\mathcal I}_{\Psi,\xi}$. For any of the six schema defined in Figure~\ref{fig:model-example} the ${\mathcal I}_{\Psi,\xi}$ set contains the same ten elements, ${\mathcal I}_{\Psi,\xi}=${\small\tt\{xcoord($v_1$), xcoord($v_2$), ycoord($v_1$), ycoord($v_2$), q0(), q1(), next($v_1$,$v_1$), next($v_1$,$v_2$), next($v_2$,$v_1$), next($v_2$,$v_2$))\}}.

Despite any element from ${\mathcal I}_{\Psi,\xi}$ can {\em a priori} appear in the $pre(\xi)$, $del(\xi)$ and $add(\xi)$ of a schema $\xi$, the space of possible \strips\ schema is constrained by the set ${\mathcal C}$ that includes: 
\begin{itemize}
\item {\em Syntactic constraints}. \strips\ constraints require negative effects appearing as preconditions, negative effects cannot be positive effects at the same time and also, positive effects cannot appear as preconditions. Formally, $del(\xi)\subseteq pre(\xi)$, $del(\xi)\cap add(\xi)=\emptyset$ and $pre(\xi)\cap add(\xi)=\emptyset$). Considering exclusively these syntactic constraints, the size of the space of possible \strips\ schema is given by $2^{2\times|{\mathcal I}_{\Psi,\xi}|}$. For every action schema in the navigation model of Figure~\ref{fig:model-example} then $2^{2\times 10}=1,048,576$.
\item {\em Domain-specific constraints}. One can also introduce domain-specific knowledge to more precisely constrain the space of possible schema. For instance in a {\em robot navigation} model like the one in Figure~\ref{fig:model-example}, {\small\tt q0()} and {\small\tt q1()} are exclusive so they cannot hold at the same time in a $pre(\xi)$/$del(\xi)$/$add(\xi)$ list. Further, {\small\tt next($v_1$,$v_1$)} and {\small\tt next($v_2$,$v_2$)} will not appear at any of these lists because the {\tt\small next} predicate is coding the {\em successor} function for {\em natural numbers}. These domain-specific constraints reduces the size of the space of possible action schema to $2^{2\times 7}=16,384$ (for every schema).
\end{itemize}

\begin{definition}[Well-defined \strips\ action schema]
Given a set of {\em predicates} $\Psi$, a list of action {\em parameters} $pars(\xi)$, and set of FOL constraints ${\mathcal C}$ we say that $\xi$ is a {\bf well-defined \strips\ action schema} iff its three lists $pre(\xi)\subseteq {\mathcal I}_{\Psi,\xi}$, $del(\xi)\subseteq{\mathcal I}_{\Psi,\xi}$ and $add(\xi)\subseteq{\mathcal I}_{\Psi,\xi}$ only contain elements in ${\mathcal I}_{\Psi,\xi}$ and they satisfy all the constraints in ${\mathcal C}$. 
\end{definition}
We say a classical planning action model is {\em well-defined} if all its corresponding  \strips\ action schema are {\em well-defined}.

\subsection{Edit distances for \strips\ action models}
First, we define two edit \emph{operations} on a schema $\xi\in\mathcal{M}$ that belongs to a \strips\ action model $\mathcal{M}\in M$:
\begin{itemize}
\item {\em Deletion}. An element is removed from any of these three lists $pre(\xi)$, $del(\xi)$ and $add(\xi)$ of the operator schema $\xi\in\mathcal{M}$ such that the resulting schema is a {\em well-defined} \strips\ action schema.
\item {\em Insertion}. An element in ${\mathcal I}_{\Psi,\xi}$ is added to any of these three lists $pre(\xi)$, $del(\xi)$ and $add(\xi)$ of the operator schema $\xi\in\mathcal{M}$ s.t. the resulting schema is {\em well-defined}.
\end{itemize}

@@

Since ${\mathcal I}_{\Psi,\xi}$ are bound sets, the maximum number of edits that can be introduced to a given action model is bound as well. The \textbf{maximum edit distance} of a \strips\ model $\mathcal{M}$ built within predicates $\Psi$ is $\delta(\mathcal{M},*)=\sum_{\xi\in\mathcal{M}} 3\times|{\mathcal I}_{\Psi,\xi}|$ (note that, if we consider the syntactic \strips\ constraints, $\delta(\mathcal{M},*)=\sum_{\xi\in\mathcal{M}} 2\times|{\mathcal I}_{\Psi,\xi}|$).

An observation $\mathcal{O}(\pi,P)$ of the execution of a plan generated with an action model $\mathcal{M}$, reflects {\em semantic knowledge} that constrains further the space of the possible schema $\xi\in \mathcal{M}$. In this case, we talk about a third kind of constraints that we call {\em observation constraints} and that can also be included into the $\mathcal{C}$ set. In addition, {\em observation constraints} allow us to define an edit distance to elicit the $P(\mathcal{O}|\mathcal{M})$ likelihoods. It can be argued that the shorter this  distance the better the given model explains the given observation.

>>>>

This paper formalizes the {\em model recognition} task where the object to recognize is not a goal but the {\em planning model} that shapes the behavior of the observed agent. Given a partially observed plan execution, and a set of possible planning models (models that share the same state variables but that update these variables with different action models); {\em model recognition} is the task of identifying the model in the set with the highest probability of producing/explaining the given observation.

@@

To better illustrate {\em model recognition}, imagine a robot in a $n\times n$ grid whose navigation is determined by the \strips\ model of Figure~\ref{fig:model-example}. According to this model, the robot can increment its {\em x-coordinate} when {\tt\small $q0$} holds (i.e. at {\em even rows} if {\tt\small $q0$} holds initially) and decrement it when {\tt\small $q1$} holds (at {\em odd rows} if {\tt\small $q0$} holds initially). Apart from this particular navigation model, different action models could be defined within the same state variables (e.g. altering the way {\tt\small $q0$} and {\tt\small $q1$} are required and updated) and these models can determine different kinds of robot navigation. Given an observation of a plan execution (like the one illustrated at Figure~\ref{fig:grid-example}) {\em model recognition} aims to identify the action model that produced/explains that observation, despite relevant information is unobserved (e.g. the applied actions or the value of state variables {\small\tt $q0$} and {\small\tt $q1$}).

{\em Model recognition} is of interest because once the planning model is recognized, then the model-based machinery for automated planning becomes applicable~\cite{ghallab2004automated}. In addition, it enables identifying different kinds of automata by observing their execution. It is well-known that diverse automata representations, like {\em finite state controllers}, {\em B\"{u}chi automata}, {\em push-down automata}, {\em {\sc GOLOG} programs} or {\em reactive policies}, can be encoded as classical planning models~\cite{baier2007exploiting,Geffner:FSM:AAAI10,patrizi2011computing,ivankovic2015optimal,segovia2017generating}.

@@

  (:action inc-y-even
    :parameters (?v1 ?v2)
    :precondition (and (ycoord ?v1) (next ?v1 ?v2) (q0))
    :effect (and (not (ycoord ?v1)) (ycoord ?v2)
                 (not (q0)) (q1)))

  (:action inc-y-odd
    :parameters (?v1 ?v2)
    :precondition (and (ycoord ?v1) (next ?v1 ?v2) (q1))
    :effect (and (not (ycoord ?v1)) (ycoord ?v2)
                 (not (q1)) (q0)))

  (:action dec-y-even
    :parameters (?v1 ?v2)
    :precondition (and (ycoord ?v1) (next ?v2 ?v1) (q0))
    :effect (and (not (ycoord ?v1)) (ycoord ?v2)
                 (not (q0)) (q1)))

  (:action dec-y-odd
    :parameters (?v1 ?v2)
    :precondition (and (ycoord ?v1) (next ?v2 ?v1) (q1))
    :effect (and (not (ycoord ?v1)) (ycoord ?v2)
                 (not (q1)) (q0)))
  \end{verbatim}           
  \end{tiny}  
 \caption{\small Example of a \strips\ action model (given in PDDL) for robot navigation in a $n\times n$ grid.}
\label{fig:model-example}
\end{figure}

@@

\subsection{The observation model}
Given a classical planning problem $P=\tup{F,A,I,G}$ and a trajectory $\tau(\pi,P)$ generated by the execution of a plan $\pi$ that solves $P$; the observation of the execution of $\pi$ on $P$ is $\mathcal{O}(\tau)=\tup{a_1^o,s_1^o \ldots , a_l^o, s_m^o}$, an interleaved combination of {\small $1\leq l\leq |\pi|$} observed actions and {\small $1\leq m\leq |\pi|$} observed states such that:
\begin{itemize}
\item Observed actions are {\em consistent} with $\pi$, which means that $\tup{a_1^o, \ldots, a_l^o}$ is a sub-sequence of $\pi$. 
\item Observed states are a sequence of {\em partial states} $\tup{s_1^o, \ldots, s_m^o}$ that is {\em consistent} with the sequence of states $\tup{s_0, s_1, \ldots, s_n}$ traversed by the trajectory $\tau(\pi,P)$. The initial state $I$ is fully observed while the observed states in $\mathcal{O}(\tau)$ may be partial states, i.e. the value of certain fluents in the intermediate states may not be observed (~$|s_i^o|\leq |F|$ for every $1\leq i\leq m$). Certain states may even be fully unobserved ($0\leq|s_i^o|$ for every $1\leq i\leq m$).  
\end{itemize}

With this regard, we are assuming that a bijective monotone mapping exists between plans and observations~\cite{ramirez2009plan}. We can then  conclude that transiting between two consecutive observed states in $\mathcal{O}(\tau)$ may require the execution of more than a single action ($\theta(s_i^o,\tup{a_1,\ldots,a_k})=s_{i+1}^o$, where ${\small k\geq 1}$ is unknown but finite). In other words, having $\mathcal{O}(\tau)$ does not imply knowing the actual length of $\pi$.

Figure~\ref{fig:grid-example} illustrates the {\em partial observation} of a six-state trajectory \{{\tt\scriptsize<(xcoord 2)(ycoord 1)>, <(xcoord 3)(ycoord 1)>, <(xcoord 4)(ycoord 1)>, <(xcoord 5)(ycoord 1)>, <(xcoord 5)(ycoord 2)>, <(xcoord 4)(ycoord 2)>}\}. In more detail, this observation only contains fluents of the kind {\tt\small (xcoord ?v)} and {\tt\small (ycoord ?v)} because the value of the remaining fluents, namely {\tt\small (next ?v1 ?v2)}, {\tt\small (q0)} and {\tt\small (q1)}, cannot be observed.

\begin{definition}[$\Phi$-observation]
Given a subset of fluents $\Phi\subseteq F$ we say that $\mathcal{O}(\tau)$ is a $\Phi$-observation of the execution of $\pi$ on $P$ iff, for every ${\small 1\leq i\leq m}$, each observed state $s_i^o$ only contains fluents in $\Phi$.
\end{definition}

\section{Model Recognition}
\label{sec:recognition}
The {\em model recognition} task is a tuple $\tup{P,M,\mathcal{O}}$ where:
\begin{itemize}
\item $P=\tup{F,A[\cdot],I,G}$ is a {\em classical planning problem} s.t. the semantics of actions $a\in A[\cdot]$ is unknown (i.e. for each $a\in A[\cdot]$ its corresponding $\rho$ and/or $\theta$ functions are undefined).
\item $M=\{\mathcal{M}_1,\ldots,\mathcal{M}_k\}$ is a non-empty {\em set of k models} for the actions in $A[\cdot]$ s.t., each model in $\mathcal{M}\in M$, defines different function pairs $\tup{\rho,\theta}$ over the state variables $F$.
\item $\mathcal{O}(\tau)$ is an {\em observation} of the execution of an unknown solution plan $\pi$ for the planning problem $P$.
\end{itemize}

{\em Model recognition} can be understood as a {\em classification task} where each class is represented with a different planning model $\mathcal{M}\in M$ and the observed plan execution $\mathcal{O}(\tau)$, is the single example to classify. The planning model associated to each class acts as the corresponding {\em class prototype} and it summarizes any observation of a plan execution that could be synthesized with that model (i.e. the set of all the examples that belong to that class).

We follow the {\em naive Bayes classifier} to assign a model $\mathcal{M}\in M$ to the given observation $\mathcal{O}(\tau)$. Therefore the {\em solution} to the {\em model recognition} task is the subset of models in $M$ that maximizes this expression.
\begin{align}
argmax_{\mathcal{M}\in M} P(\mathcal{O}|\mathcal{M}) P(\mathcal{M}).
\end{align}

\subsection{Formulating the $P(\mathcal{O}|\mathcal{M})$ {\em likelihood}}
The $P(\mathcal{M})$ probability expresses whether one model is known to be a priori more likely than the others. If this probability is not given as input it is reasonable to assume that, {\em a priori}, all models are equiprobable. The challenge in our formulation for {\em model recognition} is the definition of the $P(\mathcal{O}|\mathcal{M})$ {\em likelihood} that expresses the probability of observing $\mathcal{O}(\tau)$ when $\mathcal{M}$ is the planning model.

\begin{figure}
  \begin{scriptsize}
  \begin{center}
	\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=2cm,semithick]
	  \node[state] (A)              {$\mathcal{M}$};
	  \node[state] (B) [right of=A] {$\mathcal{M}'$};
	  \node[state] (C) [right of=B] {$\tau$};
	  \node[state] (D) [right of=C] {$\mathcal{O}$};          
	  \path
(A) edge  node {} (B)
(B) edge  node {} (C)
(C) edge  node {} (D)
;
	\end{tikzpicture}
  \end{center}
  \end{scriptsize}  
 \caption{\small {\em Bayesian network} abstracting that model $\mathcal{M}$ can be {\em transformed} into a model $\mathcal{M}'$ that is able to produce a trajectory $\tau(\pi,P)$ that reaches the goals in $P$ and is consistent with the observation $\mathcal{O}(\tau)$.}
\label{fig:net}
\end{figure}

Our approach to formulate the $P(\mathcal{O}|\mathcal{M})$ {\em likelihood} is to assess the cost of {\em transforming} $\mathcal{M}$ into a model $\mathcal{M'}$ such that: $\mathcal{M'}$ that can produce a trajectory $\tau(\pi,P)$ that reaches the goals in $P$ and is consistent with the observation $\mathcal{O}(\tau)$. Figure~\ref{fig:net} shows the four-variable {\em Bayesian network} abstracting this process. Regarding this network we have the following formulation:
\begin{align}
 P(\mathcal{O}|\mathcal{M})=\sum_{\mathcal{M}'}\sum_{\tau} P(\mathcal{M'}|\mathcal{M})P(\tau|\mathcal{M'})P(\mathcal{O}|\tau),
\end{align}
where $\tau$ ranges over all the trajectories that can be synthesized with a model $\mathcal{M}'$ and $\mathcal{M}'$ ranges over all the models that can be generated {\em transforming} $\mathcal{M}$.

Following this equation, the exact computation of $P(\mathcal{O}|\mathcal{M})$ is intractable. For most planning problems the set of trajectories consistent with an arbitrary observation can easily be huge (infinite in the case of planning problems without dead-ends)~\cite{lesh1995sound}. Even worse, the number of models $\mathcal{M}'$ that can be generated {\em transforming} a given classical planning model $\mathcal{M}$ explodes combinatorially. Instead, our approach is to formulate an effectively estimate of $P(\mathcal{O}|\mathcal{M})$ using the {\em edit distance} referred to \strips\ planning models. {\em Edit distances} are similarity metrics, traditionally computed over {\em strings} or {\em graphs}, and that has been proved successful for {\em pattern recognition}~\cite{masek1980faster,bunke1997relation}. In this work, we assess the cost of {\em transforming} a classical planning model $\mathcal{M}$ into a model $\mathcal{M'}$ formalizing and computing {\em edit distances} that are referred to \strips\ planning models.

\section{Recognition of \strips\ models}
\label{sec:asPlanning}
Here we analyze the particular instantiation of the {\em model recognition} task where the semantics of actions (i.e. $\rho$ and $\theta$ functions) are specified with \strips\ action schemata. We start formalizing the \strips\ schemata and define the full space of possible \strips\ schemata. Eventually, we introduce an {\em edit distance} for \strips\ schemata to estimate the $P(\mathcal{O}|\mathcal{M})$ likelihoods for classical planning models.

\subsection{Well-defined \strips\ action schemata}
\strips\ action schemata provide a compact representation for specifying classical planning models. Figure~\ref{fig:model-example} shows six \strips\ action schemata, codded in PDDL, that determine a particular kind of robot navigation in $n\times n$ grids.

{\em A \strips\ action schemata} $\xi$ is defined by a list of {\em parameters} $pars(\xi)$, and three list of predicates (namely $pre(\xi)$, $del(\xi)$ and $add(\xi)$) that shape the kind of fluents that can appear in the {\em preconditions}, {\em negative effects} and {\em positive effects} of the actions induced from that schemata.

We say that two \strips\ schemata $\xi$ and $\xi'$ are {\em comparable} iff $pars(\xi)=pars(\xi')$, both share the same list of parameters. For instance, we claim that the six action schemata of Figure~\ref{fig:model-example} are {\em comparable} while, for example, the {\small\tt stack(?v1,?v2)} and {\small\tt pickup(?v1)} schemata from the four operator {\em blocksworld}~\cite{slaney2001blocks} are not. Last but not least, we say that two \strips\ models $\mathcal{M}$ and $\mathcal{M}'$ are {\em comparable} iff there exists a bijective function $\mathcal{M} \mapsto \mathcal{M}^*$ that maps every action schemata $\xi\in\mathcal{M}$ to a comparable schemata $\xi'\in\mathcal{M'}$ and vice versa.

Given a \strips\ action schemata $\xi$ and a set of {\em predicates} $\Psi$ that shape the propositional state variables. The set of FOL interpretations of $\Psi$ over the corresponding parameters of the schemata $\xi$, confines the elements that can appear in the $pre(\xi)$, $del(\xi)$ and $add(\xi)$ lists. We denote this set of FOL interpretations as ${\mathcal I}_{\Psi,\xi}$. For any of the six action schemata defined in Figure~\ref{fig:model-example} the ${\mathcal I}_{\Psi,\xi}$ set contains the same ten elements, ${\mathcal I}_{\Psi,\xi}=${\small\tt\{xcoord($v_1$), xcoord($v_2$), ycoord($v_1$), ycoord($v_2$), q0(), q1(), next($v_1$,$v_1$), next($v_1$,$v_2$), next($v_2$,$v_1$), next($v_2$,$v_2$))\}}.

Despite any element from ${\mathcal I}_{\Psi,\xi}$ can {\em a priori} appear in the $pre(\xi)$, $del(\xi)$ and $add(\xi)$ of a schemata $\xi$, the space of possible \strips\ schemata is constrained by the set ${\mathcal C}$ that includes: 
\begin{itemize}
\item {\em Syntactic constraints}. \strips\ constraints require negative effects appearing as preconditions, negative effects cannot be positive effects at the same time and also, positive effects cannot appear as preconditions. Formally, $del(\xi)\subseteq pre(\xi)$, $del(\xi)\cap add(\xi)=\emptyset$ and $pre(\xi)\cap add(\xi)=\emptyset$). Considering exclusively these syntactic constraints, the size of the space of possible \strips\ schemata is given by $2^{2\times|{\mathcal I}_{\Psi,\xi}|}$. For every action schemata in the navigation model of Figure~\ref{fig:model-example} then $2^{2\times 10}=1,048,576$.
\item {\em Domain-specific constraints}. One can also introduce domain-specific knowledge to more precisely constrain the space of possible schemata. For instance in a {\em robot navigation} model like the one in Figure~\ref{fig:model-example}, {\small\tt q0()} and {\small\tt q1()} are exclusive so they cannot hold at the same time in a $pre(\xi)$/$del(\xi)$/$add(\xi)$ list. Further, {\small\tt next($v_1$,$v_1$)} and {\small\tt next($v_2$,$v_2$)} will not appear at any of these lists because the {\tt\small next} predicate is coding the {\em successor} function for {\em natural numbers}. These domain-specific constraints reduce further the size of the space of possible action schemata to $2^{2\times 7}=16,384$ (for every schemata in the navigation model of Figure~\ref{fig:model-example}).
\end{itemize}

\begin{definition}[Well-defined \strips\ action schemata]
Given a set of {\em predicates} $\Psi$, a list of action {\em parameters} $pars(\xi)$, and set of FOL constraints ${\mathcal C}$ we say that $\xi$ is a {\bf well-defined \strips\ action schemata} iff its three lists $pre(\xi)\subseteq {\mathcal I}_{\Psi,\xi}$, $del(\xi)\subseteq{\mathcal I}_{\Psi,\xi}$ and $add(\xi)\subseteq{\mathcal I}_{\Psi,\xi}$ only contain elements in ${\mathcal I}_{\Psi,\xi}$ and they satisfy all the constraints in ${\mathcal C}$. 
\end{definition}
We say a classical planning action model is {\em well-defined} if all its corresponding  \strips\ action schemata are {\em well-defined}.

\subsection{Edit distances for \strips\ action models}
First, we define the two edit \emph{operations} on a schemata $\xi\in\mathcal{M}$ that belongs to a \strips\ action model $\mathcal{M}\in M$:
\begin{itemize}
\item {\em Deletion}. An element is removed from any of these three lists $pre(\xi)$, $del(\xi)$ and $add(\xi)$ of the operator schemata $\xi\in\mathcal{M}$ such that the resulting schemata is a {\em well-defined} \strips\ action schemata.
\item {\em Insertion}. An element in ${\mathcal I}_{\Psi,\xi}$ is added to any of these three lists $pre(\xi)$, $del(\xi)$ and $add(\xi)$ of the operator schemata $\xi\in\mathcal{M}$ s.t. the resulting schemata is {\em well-defined}.
\end{itemize}

@@

Since ${\mathcal I}_{\Psi,\xi}$ are bounded sets, the maximum number of edits that can be introduced to a given action model is bounded as well. The \textbf{maximum edit distance} of a \strips\ model $\mathcal{M}$ built within predicates $\Psi$ is $\delta(\mathcal{M},*)=\sum_{\xi\in\mathcal{M}} 3\times|{\mathcal I}_{\Psi,\xi}|$ (note that, if we consider the syntactic \strips\ constraints, $\delta(\mathcal{M},*)=\sum_{\xi\in\mathcal{M}} 2\times|{\mathcal I}_{\Psi,\xi}|$).

An observation $\mathcal{O}(\tau)$ of the execution of a plan generated with an action model $\mathcal{M}$, reflects {\em semantic knowledge} that constrains further the space of the possible schemata $\xi\in \mathcal{M}$. We talk then about a third kind of constraints, that we call {\em observation constraints}, and that can also be included into the $\mathcal{C}$ set. In addition, {\em observation constraints} allow us to define an edit distance to elicit the $P(\mathcal{O}|\mathcal{M})$ likelihoods. It can be argued that the shorter this  distance the better the given model explains the given observation.
